Automatic word stress annotation of Russian running text (max 8 pages)

### 2015-01-05 rusaccent-nodalida-2015/rusaccent.tex
more
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@58230 72bbbca6-d526-0410-a7d9-f06f51895060

\title{Automatic accent annotation for Russian}

@@

\date{}

\begin{document}
\maketitle
\begin{abstract}
  This document contains the instructions for preparing a camera-ready
  manuscript for the proceedings of Nodalida-2015. The document itself
  conforms to its own specifications, and is therefore an example of
  what your manuscript should look like. These instructions should be
  used for both papers submitted for review and for final versions of
  accepted papers.  Authors are asked to conform to all the directions
  reported in this document.
\end{abstract}

@@

Russian word stress is variable, which means that the stress position in a given word is not 
generally predictable. Furthermore, the inflecting word classes include complex patterns of
shifting stress. Lexical stress and its attendant vowel reduction are a prominent feature of
spoken Russian, but standard written Russian does not typically mark word stress.\footnote{
Texts intended for native speakers sometimes mark stress on words that cannot be 
disambiguated through context.} Without information about lexical stress position, 
correctly converting written Russian text to speech is impossible. This is true both for humans
(e.g. foreign language students) and computers (e.g. text-to-speech or automatic speech 
recognition).

@@

\begin{table}
\begin{center}
\begin{tabular}{c|ll}
\bf Number-Case & \bf `castle' & \bf `lock' \\
\hline
\texttt{SG-NOM} & \rus{з\'{a}мок} & \rus{зам\'{о}к} \\
\texttt{SG-GEN} & \rus{з\'{a}мка} & \rus{замк\'{а}} \\
\texttt{PL-NOM} & \rus{з\'{a}мки} & \rus{замк\'{и}} \\
\texttt{PL-GEN} & \rus{з\'{a}мков} & \rus{замк\'{о}в} \\
... & ... & ... \\
\end{tabular}
\end{center}
\caption{ \label{table:MScongruent} Morphosyntactically congruent homographs. }
\end{table}

@@

\ex. Morphosyntactically incongruent homographs \label{ex:MSincongruent} \rr{more interesting examples? \rus{нашей или было или дорога} }
\a. \rus{попад\'{а}ть} \emph{popadát'} `fall.\textsubscript{INF-IPFV}' \\
     \rus{поп\'{а}дать} \emph{popádat'} `fall.\textsubscript{INF-PFV}'
\b. \rus{узнаёт} \emph{uznaët} `finds out.\textsubscript{PRS-IPFV-3SG}' \\
     \rus{узн\'{а}ет} \emph{uznáet} `finds out.\textsubscript{FUT-PFV-3SG}'

\rr{another relevant issue is whether the ambiguity can be resolved at all. Maybe this is something we should bring up in the discussion section. For example, when is stress marked in authentic by-native-for-native texts? A computer should not be expected to predict the stress on such words (although a computer could possibly identify words that should be marked for natives.)}
Different types of stress position ambiguity are overcome by different information. Detailed 
part-of-speech tagging can determine the stress of intraparadigmatic homographs and
morphosyntactically incongruent homographs, whereas word sense disambiguation can 
determine the stress position of morphosyntactically congruent homographs. In the case of
running text in Russian, most stress placement ambiguity is rooted in morphosyntactic
ambiguity, so in this study we focus on the role of part-of-speech tagging in improving
automatic stress placement, leaving word sense disambiguation to future work.

@@

Many studies 
of Russian text-to-speech and automatic speech recognition make note of the 
difficulties caused by the shortcomings of their stress-marking resources (e.g. CITE Krivnova 1998).
\cite{Xomicevic_2008} developed a set of heuristics for guessing stress placement on 
unknown words in Russian. More recently, CITE Hall \& Sproat (2013) trained a maximum entropy
model on a dictionary of Russian words, achieving 83.9\% accuracy for stress on unknown words 
and 98.7\% on known words. One major shortcoming of evaluating on wordlists is 
that they do not reflect wordform frequency and opportunities for disambiguation.
Many of the most problematic stress ambiguities in Russian occur in 
high-frequency wordforms, so evaluations based on wordlists - as opposed to running
text - have inflated results. An added benefit of working with running text
is the possibility of disambiguating homographs based on syntactic context.

So far, the few studies related to automatic stress placement in Russian have been aimed at
text-to-speech and speech recognition technology. However, the target application of our stress
annotator is a different domain: language learning. Since standard Russian does not contain 
stress-markings, learners are frequently unable to pronounce unknown words correctly without
referencing a dictionary or similar resources. In the context
of language learning, marking stress incorrectly is arguably worse than not marking it
at all. Because of this, we want our stress annotator to be able to abstain from marking
stress on words that it is unable to resolve with high confidence.

@@

State-of-the-art part-of-speech and morphological analysis in Russian is based on finite-state
technology. We developed free and open-source finite-state tools with the ability to analyze and
generate stressed wordforms. A finite-state morphological analyzer generates all possible readings
of each wordform, and the constraint grammar 
to analyze running text, and to generate stressed wordforms.

Manuscripts must be in two-column format.  Exceptions to the
two-column format include the title, authors' names and complete
addresses, which must be centered at the top of the first page, and
any full-width figures or tables (see the guidelines in
Subsection~\ref{ssec:fonts}). {\bf Type single-spaced.}  Start all
pages directly under the top margin. See the guidelines later
regarding formatting the first page (Subsection~\ref{ssec:first}).
The manuscript should be printed single-sided and its length should
not exceed the maximum page limit described in
Section~\ref{sec:length}.  {\bf Do not number the pages.}

>>>>

\title{Automatic word stress annotation of Russian running text (max 8 pages)}

@@

\date{2015}

\begin{document}
\maketitle
\begin{abstract}
  We evaluate the effectiveness of new tools that we developed for automatically annotating 
  word stress in Russian running text. Given an adequate lexicon with specified stress, the
  primary obstacle for correct stress placement is disambiguating homographic forms. Using a
  Constraint Grammar, we achieve a X\% improvement over the baseline undisambiguated performance.
\end{abstract}

@@

Russian word stress is variable. The inflecting word classes include complex patterns of
shifting stress, and a lexeme's stress pattern cannot be predicted from surface forms.
Although lexical stress and its attendant vowel reduction are a prominent feature of
spoken Russian, standard written Russian does not typically mark word stress.\footnote{
Texts intended for native speakers sometimes mark stress on words that cannot be 
disambiguated through context, although this is somewhat rare.} Without information about lexical 
stress position, 
correctly converting written Russian text to speech is impossible. This is true both for humans
(e.g. foreign language students) and computers (e.g. text-to-speech).

@@

\begin{table}
\begin{center}
\begin{tabular}{c|ll}
\bf Number-Case & \bf `castle' & \bf `lock' \\
\hline
\texttt{SG-NOM} & \rus{з\'{a}мок} & \rus{зам\'{о}к} \\
\texttt{SG-GEN} & \rus{з\'{a}мка} & \rus{замк\'{а}} \\
\texttt{PL-NOM} & \rus{з\'{a}мки} & \rus{замк\'{и}} \\
\texttt{PL-GEN} & \rus{з\'{a}мков} & \rus{замк\'{о}в} \\
... & \hskip 1em ... & \hskip 1em ... \\
\end{tabular}
\end{center}
\caption{ \label{table:MScongruent} Morphosyntactically congruent homographs. }
\end{table}

@@

\ex. Morphosyntactically incongruent homographs \label{ex:MSincongruent}
\a. \rus{попад\'{а}ть} \emph{popadát'} `fall.\textsubscript{INF-IPFV}' \\
    \rus{поп\'{а}дать} \emph{popádat'} `fall.\textsubscript{INF-PFV}'
\b. \rus{узнаёт} \emph{uznaët} `finds out.\textsubscript{PRS-IPFV-3SG}' \\
    \rus{узн\'{а}ет} \emph{uznáet} `finds out.\textsubscript{FUT-PFV-3SG}'
\c. \rus{н\'{а}шей} \emph{nášej} `our.\textsubscript{F-SG-GEN/DAT/LOC...}'\\
    \rus{наш\'{е}й} \emph{našéj} `sew on.\textsubscript{IMP-2SG}'
\d. \rus{дор\'{о}га} \emph{doróga} `road.\textsubscript{N-F-SG-NOM}'\\
    \rus{дорог\'{а}} \emph{dorogá} `dear.\textsubscript{ADJ-F-SG-PRED}'

\rr{another relevant issue is whether the ambiguity can be resolved at all. Maybe this is 
something we should bring up in the discussion section. For example, when is stress marked in 
authentic by-native-for-native texts? A computer should not be expected to predict the stress on 
such words (although a computer could possibly identify words that should be marked for 
natives.)} It should be noted that not all morphosyntactic ambiguity leads to stress ambiguity. For example,
all inflecting modifiers in Russian have the same form for F-SG-GEN, F-SG-LOC, F-SG-DAT and 
F-SG-INS. We refer to this as \emph{stress-irrelevant} ambiguity, since all readings have
the same stress placement. Stress-irrelevant ambiguity can be intraparadigmatic and/or interlexical.

Different types of stress ambiguity are overcome by different information. Detailed 
part-of-speech tagging can determine the stress of intraparadigmatic homographs and
morphosyntactically incongruent homographs, whereas word sense disambiguation can 
determine the stress position of morphosyntactically congruent homographs. In the case of
running text in Russian, most stress placement ambiguity is rooted in morphosyntactic
ambiguity, so in this study we focus on the role of part-of-speech tagging in improving
automatic stress placement, leaving word sense disambiguation to future work.

@@

Many studies of Russian text-to-speech and automatic speech recognition make note of the 
difficulties caused by the shortcomings of their stress-marking resources (e.g. CITE Krivnova 
1998), but very few studies have targeted stress-marking itself, and they tend to
focus on place stress on unknown words, with almost no attention to the problems that
arise with known words. For example, 
\cite{Xomicevic_2008} developed a set of heuristics for guessing stress placement on 
unknown words in Russian. More recently, CITE Hall \& Sproat (2013) trained a maximum entropy
model on a dictionary of Russian words, achieving 83.9\% accuracy for stress on unknown words 
and 98.7\% on known words. One major shortcoming of evaluating on wordlists is 
that they do not reflect wordform frequency.
Many of the most problematic stress ambiguities in Russian occur in 
high-frequency wordforms\rr{might it be worth computing the difference in stress
ambiguity between wordform lists and running text? i.e. what percentage of tokens have ambiguous
stress.}, so evaluations based on wordlists - as opposed to running
text - have inflated results. Furthermore, working with running text allows for the
possibility of disambiguating homographs based on syntactic context.

So far, the implicit target application the few studies related to automatic stress placement
in Russian has been text-to-speech and speech recognition technology. However, the target 
application of our stress
annotator is a different domain: language learning. Since standard Russian does not contain 
stress-markings, learners are frequently unable to pronounce unknown words correctly without
referencing a dictionary or similar resources. In the context
of language learning, marking stress incorrectly is arguably worse than not marking it
at all. Because of this, we want our stress annotator to be able to abstain from marking
stress on words that it is unable to resolve with high confidence.

@@

State-of-the-art part-of-speech and morphological analysis in Russian is based on finite-state
technology.\footnote{add footnote or citations for Dialog, mystem, etc.} We developed free
and open-source finite-state tools capable of analyzing and
generating stressed wordforms. A finite-state transducer\footnote{a two-level 
morphology, xfst/hfst, blah, blah, blah} generates all possible readings
of each wordform, and a constraint grammar\footnote{vislcg3, blah blah} then removes
some extra readings based on syntactic context. The ultimate success of our stress placement
system depends on the performance of the Constraint Grammar. Ideally, the Constraint Grammar would
successfully remove all but the correct reading for each token, but in practice some
tokens still have more than one reading remaining. Therefore,
we also evaluate various approaches to deal with the remaining ambiguity,
as described in the following paragraph and Table~\ref{tab:conditions}.

\begin{table*}[t]
  \centering
  \begin{tabular}{r|c|c}
    & \rus{кость}-N-F-SG-GEN\hskip 1em\rus{к\'{о}сти} & \rus{кость}-N-F-SG-GEN\hskip 1em\rus{к\'{о}сти} \\
    & \rus{кость}-N-F-SG-DAT\hskip 1em\rus{к\'{о}сти} & \rus{кость}-N-F-SG-DAT\hskip 1em\rus{к\'{о}сти} \\
    & \rus{костить}-V-IPFV-IMP\hskip 1em\rus{кост\'{и}} & \\
    \hline
    {\small {\tt bare}} & \rus{кости} & \rus{кости} \\
    {\small {\tt safe}} & \rus{кости} & \rus{к\'{о}сти} \\
    {\small {\tt guess}} & \rus{к\'{о}сти} or \rus{кост\'{и}} & \rus{к\'{о}сти}\\
    {\small {\tt guessFreq}} & \rus{к\'{о}сти} & \rus{к\'{о}сти}
  \end{tabular}
  \caption{Example output of each condition, given a particular set of readings}
  \label{tab:conditions}
\end{table*}

The {\small {\tt bare}} approach is to not mark stress on words with more than one reading.
The {\small {\tt safe}} approach is to mark stress only on
tokens whose morphosyntactic ambiguity is stress-irrelevant. The {\small {\tt guess}}
approach is to randomly select one of the available readings.
The {\small {\tt guessFreq}} approach is to select the reading that is most frequent
in our corpus. If none of the readings is found in the corpus, then it selects the reading
with the tagset that is most frequent in our corpus. Note that for tokens with
stress-irrelevant ambiguity, {\small {\tt guess}} and {\small {\tt guessFreq}} produce
the same result as the {\small {\tt safe}} method.

For our baseline, we take the output of our morphological analyzer (without the Constraint 
Grammar) in combination with {\small {\tt bare}}, {\small {\tt safe}}, {\small {\tt guess}} and {\small {\tt guessFreq}}. The purpose of this study is to compare the
